pema application

### 2022-04-20 paper/manuscript.Rmd
simplify applied example

```{r setup, include = FALSE}
library("papaja")
library(tidySEM)
library(kableExtra)
r_refs("r-references.bib")
out <- readRDS("output.RData")
```

@@

>>>>

```{r setup, include = FALSE}
run_everything = FALSE
library("papaja")
library(tidySEM)
library(kableExtra)
r_refs("r-references.bib")
out <- readRDS("output.RData")
```

@@

# Applied example

In this application, we will work with the `pema::bonapersona` data [@BonapersonaEtal2019].
This meta-analysis of over 400 experiments investigated the effects of early life adversity on cognitive performance in rodents.
This example uses a small subset of the more than 30 moderators.
<!-- We focus specifically on how to choose the different hyperparameters of the regularized horseshoe prior. -->

```{r exdata, eval = TRUE}
# Set random seed
set.seed(1)
# Load relevant packages
library(pema)
library(tidySEM)
library(ggplot2)
library(bayesplot)

# Select data to analyze
df <- bonapersona[ , c("yi", "vi", "mTimeLength", "year", "model", "ageWeek")] 
# Impute missings
df$ageWeek[is.na(df$ageWeek)] <- median(df$ageWeek, na.rm = TRUE)
```

## Visualizing the prior

Our simulation study shows good performance with default hyperparameters.
However, experienced users may want to customize the prior.
Visualizing the prior can be helpful in this process.
This is accomplished using the interactive application visualization application available through `shiny_prior()`.
The user can plot the prior distributions resulting from different sets of hyperparameters and compare them.
Increasing the values of the scale parameters (`scale_global` and `hs_scale_slab`) results in a more spread out prior, which applies less regularization.
Increasing the degrees of freedom (`df_global` and `df_slab`) results in thinner tails, which applies more regularization.
<!-- One advantage of the regularized horseshoe prior is its ability to define the prior in terms of the expected number of nonzero coefficients, using the argument `relevant_pars`. -->
<!-- This argument in turn influences the other hyperparameters. -->
<!-- We can also plot the implied prior on the effective number of nonzero coefficients given specific hyperparameters. -->
<!-- In this application, we have a total of 7 possible moderators, including the dummy variables for the categorical moderator "model". We can visualize the implied prior on the number of nonzero coefficients for different values of `relevant_pars`: p0 = 1; p0 = 3; p0 = 6; or p0 = NA, in which case no prior information is given and `scale_global`= 1. -->

```{r exprior, eval = FALSE, include = FALSE}
mod <- stan_model("../inst/stan/prior_meff.stan")
datdum <- data.frame(model.matrix(~., df)[, -1])
meff_prior_fun <- function(priorinfo = TRUE, p0){
  D <- ncol(datdum) - 2
  n <-  nrow(datdum)
  sigma <- 1

  vars <- apply(datdum, 2, var)[-c(1:2)]

  if(priorinfo == TRUE){
    tausq0 <-  p0/(D-p0)*(sigma/sqrt(n))
  } else if(priorinfo == FALSE){
    p0 <- NA
    tausq0 <- 1
  }

  dat <- list(tausq0 = tausq0,
              D = D,
              n = n,
              sigma = sigma, # fixed to 1 in the current implementation; alternative would be to scale to tau
              s2 = vars)
  fit <- sampling(mod, data = dat, iter = 4000)
  draws <- as.matrix(fit)
  df <- cbind.data.frame("p0" = paste0("p0 = ", p0),
                         "draws" = draws[, "meff"])
  return(df)
}

df1 <- meff_prior_fun(priorinfo = FALSE)
df2 <- meff_prior_fun(priorinfo = TRUE, p0 = 1)
df3 <- meff_prior_fun(priorinfo = TRUE, p0 = 3)
df4 <- meff_prior_fun(priorinfo = TRUE, p0 = 6)
df <- rbind.data.frame(df1, df2, df3, df4)

# plot
ggplot(df, aes(x=draws)) + 
  geom_histogram() +
  facet_wrap(~p0) +
 theme_bw() + 
  xlab("Effective number of nonzero coefficients") + 
  ylab("") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
<!-- It is clear that the implied prior on the number of nonzero coefficients follows the prior information it has been given, with more mass on a small number of relevant moderators if that is expected a priori (p0 = 1) and more mass on a larger number of relevant moderators if that is expected a priori (p0 = 6). Interestingly, when no prior information is provided (p0 = NA), the default `global_scale` setting of 1 implies a relatively large number of nonzero coefficients, which is important to keep in mind when using this default setting. -->

In this example, we will estimate the model using default settings, which includes a horseshoe prior with default hyperparameters.
To see the default values, open the function documentation using `?brma`.

```{r exmodel1, echo = TRUE, eval = FALSE}
fit_hs1 <- brma(yi ~ ., data = df, vi = "vi")
```
```{r, echo = FALSE, eval = run_everything}
fit_hs1 <- brma(yi ~ ., data = df, vi = "vi")
saveRDS(fit_hs1, "fit_hs1.RData")
```
```{r, echo = FALSE, eval = TRUE}
fit_hs1 <- readRDS("fit_hs1.RData")
```

Note the warning message about divergent transitions, which may indicate problems with model convergence.
This warning is generated by `Stan`, the computational back-end of `brma()`.
Divergent transitions can result in biased estimates. 
However, if the number of divergences is small and there are no further indications of non-convergence,
the posterior distribution is often good enough to safely interpret the results.
In some cases, this problem may be resolved by increasing the values for the degrees of freedom parameters.
Note that increasing both `df_global` and `df_slab` to 5 results in fewer divergences,
but does not otherwise influence the substantive interpretation of the results:

```{r exmodel2, eval = FALSE, echo = TRUE}
fit_hs2 <- brma(yi ~ .,
                data = df,
                vi = "vi",
                prior = c(df = 5, df_global = 5, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = NULL))
```
```{r, echo = FALSE, eval = run_everything}
fit_hs2 <- brma(yi ~ .,
                data = df,
                vi = "vi",
                prior = c(df = 5, df_global = 5, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = NULL))
saveRDS(fit_hs2, "fit_hs2.RData")
```
```{r, echo = FALSE, eval = TRUE}
fit_hs2 <- readRDS("fit_hs2.RData")
```

In general, it is prudent to assess model convergence.
Various numeric and visual diagnostics exist, for example, in the R-package `rstan`.
These functions can be used by converting a `brma` object to a `stanfit` object, using the function `pema::as.stan()`:

```{r}
fit_stan <- as.stan(fit_hs1)
check_hmc_diagnostics(fit_stan)
```

Additionally, traceplots can be created for each parameter to visualize the MCMC draws using the `bayesplot` R-package.
The traceplots of a converged model look like "fat caterpillars", with the different MCMC chains mixing together: 
<!-- Note that there are many auxiliary parameters used for the regularized horseshoe prior. No information on these parameters is returned in the `brma` fitobject, however, once the fitobject is transformed into a Stan object for plotting, these parameters are included. -->

```{r extrace, eval = FALSE, echo = TRUE}
mcmc_trace(fit_stan, pars = c("Intercept", "year"))
```
```{r, eval = run_everything, echo = FALSE}
p <- mcmc_trace(fit_stan, pars = c("Intercept", "year"))
ggsave("traceplot.png", p, device = "png", width = 4, height = 3, dpi = 300, scale = 1.5)
```
```{r, eval = TRUE, echo = FALSE}
knitr::include_graphics("traceplot.png")
```

Convergence can also be checked by examining for the summary of the model object.
The column `Rhat` provides information on the split $\hat{R}$, a version of the potential scale reduction factor (PSRF; @GelmanRubin1992).
Values close to 1 indicate convergence.
In addition, the column `n_eff` provides information on the number of effective (independent) MCMC samples, which should be high relative to the total number of samples (in this case, 4000):

```{r exsumm, eval = TRUE}
summary(fit_hs1)
```
All `Rhat` values are close to 1, indicating convergence.
The effective number of MCMC samples is relatively small compared to the total number of MCMC samples.
An often used heuristic is to consider ratios smaller than 0.1 as problematic.
Even though this is not the case here, consider increasing the number of MCMC iterations.
<!-- These diagnostics can be easily visualized using the `bayesplot` package: -->

```{r exrhat, eval = FALSE, echo = FALSE}
rhats <- rhat(fit)
mcmc_rhat(rhats[names(rhats) %in% parsel])
```

```{r exeffn, eval = FALSE, echo = FALSE}
ratios_neff <- neff_ratio(fit)
mcmc_neff(ratios_neff[names(ratios_neff) %in% parsel])
```

<!-- ### Results -->

The `summary()` function also provides summary statistics for the model parameters.
For each parameter, the posterior mean, standard deviation, and quantiles are reported by default (the 50% quantile is the median).
To perform inference, consider using the posterior median and 95% credible interval (2.5% - 97.5%).
Since Bayesian penalization does not automatically shrink estimates exactly to zero,
we retained moderators when the 95% credible interval excluded zero in the simulation study.
The `summary()` flags such parameters with an asterisk.
If the model is converted using `as.stan()`, the results can also be visualized using standard `rstan` functions.
In this example, we conclude that there are no moderators for which the 95% CI excludes zero.

```{r exres, eval = FALSE}
plot(fit_stan, pars = rownames(coef(fit_hs1))[2:8])
```
```{r, eval = run_everything, echo = FALSE}
p <- plot(fit_stan, pars = rownames(coef(fit_hs1))[2:8])
ggsave("plot_es.png", p, device = "png", width = 4, height = 3, dpi = 300, scale = 1.5)
```
```{r, eval = TRUE, echo = FALSE}
knitr::include_graphics("plot_es.png")
```

<!-- ## Prior sensitivity -->

<!-- To assess the sensitivity of the results to the chosen values of the hyperparameters, -->
<!-- one can run the model with a different prior. -->
<!-- In this example, we vary the expected number of relevant moderators `relevant_pars`.  -->

```{r expriorsens, eval = FALSE, echo = FALSE}
fit_hs3 <- brma(yi ~ .,
                data = datdum,
                vi = "vi",
                method = "hs",
                standardize = FALSE,
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = 1),
                mute_stan = FALSE)

fit_hs4 <- brma(yi ~ .,
                data = datdum,
                vi = "vi",
                method = "hs",
                standardize = FALSE,
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = 3),
                mute_stan = FALSE)

fit_hs5 <- brma(yi ~ .,
                data = datdum,
                vi = "vi",
                method = "hs",
                standardize = FALSE,
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = 6),
                mute_stan = FALSE)

# plot
df.fun <- function(fit, prior){
  plotdat <- data.frame(fit$coefficients)
  plotdat$par <- rownames(plotdat)
  plotdat$Prior <- prior
  return(plotdat)
}

df1 <- df.fun(fit_hs1, prior = "hs default")
df3 <- df.fun(fit_hs3, prior = "hs with p0 = 1")
df4 <- df.fun(fit_hs4, prior = "hs with p0 = 3")
df5 <- df.fun(fit_hs5, prior = "hs with p0 = 6")

df <- rbind.data.frame(df1, df3, df4, df5)
df$par <- as.factor(df$par)
plotdat <- filter(df, !par %in% c("Intercept", "tau2"))

pd <- 0.5

ggplot(plotdat, aes(x=mean, y=par, group = Prior)) + 
  geom_errorbar(aes(xmin=X2.5., xmax=X97.5., colour = Prior), width=.1, position = position_dodge(width = pd)) +
  geom_point(aes(colour = Prior), position = position_dodge(width = pd)) +
  theme_bw() + xlab("Posterior mean") + ylab("")
```

<!-- When running the models with prior information regarding the number of relevant moderators, it becomes clear that there are convergence issues. The number of divergent transitions is very high and the maximum treedepth is often exceeded. For the model with `relevant_pars` = 1, these convergence issues are also reflected in high `rhat` values. -->
<!-- When comparing the results, it is clear that the horseshoe specifications that include prior information shrink coefficients more towards zero compared to the default setting. This is a consequence of the fact that these settings imply more prior mass on smaller numbers of relevant moderators, as was shown previously. However, in this situation, this high shrinkage results in convergence issues so these results should not be fully trusted but are only shown for the purpose of illustration. In our experience, the default prior settings are suitable for most applications, but conducting a prior sensitivity analysis as well as prior visualizations before running the analysis can help users to better understand the prior and its influence on the results. -->
